---
title: "Lecture 16: Spatial Analysis, Part 1"
author: "Nick Hagerty* <br> ECNS 460/560 Fall 2023 <br> Montana State University"
date: ".small[<br> *Adapted from [“R Geospatial Fundamentals”](https://github.com/dlab-berkeley/R-Geospatial-Fundamentals) by the UC Berkeley D-Lab, used under [CC BY-NC 4.0](https://creativecommons.org/licenses/by-nc/4.0/).]"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
      fig_caption: true
---
name: toc

```{css, echo=FALSE}
# CSS for including pauses in printed PDF output (see bottom of lecture)
@media print {
  .has-continuation {
    display: block !important;
  }
}
.remark-code-line {
  font-size: 95%;
}
.small {
  font-size: 75%;
}
.medsmall {
  font-size: 90%;
}
.scroll-output-full {
  height: 90%;
  overflow-y: scroll;
}
.scroll-output-75 {
  height: 75%;
  overflow-y: scroll;
}
```

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(knitr)
library(tidyverse)
knitr::opts_chunk$set(
	fig.align = "center",
	cache = TRUE,
	dpi = 300,
  warning = F,
  message = F
)
```

# Table of contents

**Part 1**

1. [Spatial data and quick mapping](#start)

1. [Reference systems and projections](#crs)

**Part 2**

1. Spatial queries

1. Spatial subsetting

1. Geometry operations

1. Spatial joins


---
class: inverse, middle
name: start

# Spatial data and quick mapping

---

# GIS

The most widespread **geographic information system** is **ArcGIS.**

**Advantages of ArcGIS:**
* Avoid coding.
* Interface for browsing and exploring data is incredibly comprehensive and fast.

**Why we're using R instead:**
- Free.
- Reproducible.
- Scriptable.
- Easily integrated with the rest of your project.
- Easy to export attractive, professional maps.
- Honestly, easier if you know some R already.

---

# sf: Simple Features

`sf` is the main package for working with **vector** data in R.

Install and load it (and a couple others):
```{r}
library(sf)
library(tidyverse)
library(tmap)
```

Set your working directory (in your console) to the location of this file. E.g.:
```{r, eval=F}
setwd("C:/git/491/course-materials/lecture-slides/10-Spatial")
```

---

# Shapefiles

The ESRI **shapefile** is the most widely used type of file format for storing geospatial vector data. A "shapefile" is actually a collection of 3+ files:

Required files:
- `shp`: The main file that stores the feature geometry
- `shx`: A positional index for locating the feature geometry in the `shp` file
- `dbf`: The data table (in dBase IV format) that stores the attribute information for each feature

Optional files:
- `prj`: Stores the coordinate reference system information. (**should be required!**)
- `sbn` and `sbx`: spatial index to speed up geometry operations (*used only by ESRI software*)
- `xml`: Metadata — Stores information about the shapefile.
- `cpg`: Specifies the code page for identifying the character encoding set to be used.

All files need to be kept together in the same directory.

---

# Load shapefile data

List the files:
```{r}
dir("data/california_counties")
```

Load the data:
```{r}
counties = st_read("data/california_counties/CaliforniaCounties.shp")
```

---

# Geodatabases

Shapefiles have some severe limitations.
* They must be less than 2 GB.
* Column names cannot be longer than 10 characters.
* The number of columns is limited to 255.

Another, newer file format is called a **geodatabase.** 

`st_read` can handle geodatabases with the `layer` argument.
* The important thing to keep in mind is that in your computer, the `.gdb` file *appears* to be a folder, but the individual files within it are uninterpretable.
* `st_layers` will show you the list of layers in a geodatabase.


---

# Quick mapping

With base R:
```{r, out.width="55%"}
plot(counties["MED_AGE"])
```

---

# Quick mapping

With tmap:
```{r, out.width="55%"}
qtm(counties)
```

---

# Quick mapping

With tmap, interactively:
```{r, error = T}
tmap_mode("view")   # the default is mode = "plot"
qtm(counties)
```

--

This is a common problem and only takes a quick fix:
```{r}
counties = st_read("data/california_counties/CaliforniaCounties.shp") |>
  st_make_valid()
```

---

# Quick mapping

With tmap, interactively:
```{r, out.height="75%", out.width="100%"}
tmap_mode("view")   # the default is mode = "plot"
qtm(counties)
```

---

# sf objects

An **sf** object is a standard data frame with an extra **geometry** column.
.scroll-output-full[
```{r}
str(counties)
```
]

---

# sf objects

The geometry column is **sticky:**
```{r}
head(counties[1:3])
```

---

# sf objects

If you want to remove the geometry:
```{r}
counties_df = st_drop_geometry(counties)
head(counties_df[1:3])
class(counties_df)
```

---

# Data wrangling works normally

Map all counties and overlay Alameda County in green:
```{r, out.height="70%", out.width="100%"}
alameda = counties |> filter(NAME == "Alameda")
map_alameda = tm_shape(counties) + tm_polygons(border.col = "white") +
  tm_shape(alameda) + tm_borders(col = "green", lwd = 3)
map_alameda
```

---

# Save your map and your shapefile

```{r}
# Static image
tmap_save(map_alameda, filename = "output/map_alameda.png")

# Interactive version
tmap_save(map_alameda, filename = "output/map_alameda.html")

# Shapefile
st_write(alameda, dsn = "output/alameda.shp", delete_dsn = TRUE)
```


---
class: inverse, middle
name: crs

# Reference systems and projections

---

# Another shapefile

Load in another shapefile of US state borders:
```{r}
states = st_read("data/us_states_contiguous/states_contiguous.shp")
```

---

# Another shapefile

Plot it alongside counties:
```{r, out.width="55%"}
plot(counties$geometry, col = 'lightgrey', border = 'white')
plot(states$geometry, col = 'blue', border = 'red', lwd = 5, add = T)
```

---

# Another shapefile

What do you think happened?

--

```{r}
st_bbox(counties)
st_bbox(states)
```


---

# Getting the CRS

What is the CRS of `states`?
```{r}
st_crs(states)
```

---

# Getting the CRS

And the CRS of `counties`?
.scroll-output-full[
```{r}
st_crs(counties)
```
]

---

# EPSG codes

CRSs are most commonly referenced by EPSG codes, which you can Google.
* The counties shapefile was in California Albers Equal, or EPSG:3310. [https://epsg.io/3310](https://epsg.io/3310)

To get two shapefiles to work with each other, we need to give them the same CRS.     
To change a CRS, we need to **project** (or re-project) our data.

**Do you *have* to use a projected CRS?**
- No -- much of the time, it's fine to keep your spatial data in a geographic CRS.
- Most functions will project "on the fly" using a default.

**When *should* you use a projected CRS?**
- When doing calculations, like area or distance.
- When you want to control the appearance of your map output.

**How should you choose a projected CRS?**
- What you want to preserve (area, direction, or distance).
- Location on Earth of the area of focus.

---

# Common projections

**1. Web Mercator**
- Preserves direction/angle/shape but distorts area and distance.
- Decent starting point for most places in the world.
```{r, out.width="50%"}
states_mercator = st_transform(states, crs = 3857)
tmap_mode("plot")
qtm(states_mercator)
```

---

# Common projections

**2. U.S. National Atlas (Albers) Equal Area**
- Preserves area but distorts direction/angle/shape and distance.
- My favorite for the continental U.S.
```{r, out.width="50%"}
states_albers = st_transform(states, crs = 2163)
qtm(states_albers)
```

---

# Common projections

**3. UTM Zone 11N**
- Preserves direction/angle/shape but distorts area and distance.
- Different UTM zones are centered at different locations.
- Good for maps of smaller areas.
```{r, out.width="45%"}
states_utm11N = st_transform(states, crs = 2955)
qtm(states_utm11N)
```

---

# Common projections

**4. Pseudo Plate Caree**
- Distorts everything! Simply a graph of latitude vs. longitude.
- Common default, but no excuse for using it in 2022.
```{r, out.width="65%"}
plot(states$geometry, asp = 1)
```

---

# (Re)projecting a shapefile

The safest way to reproject is to directly reference the CRS of another layer:
```{r, out.width="55%"}
counties_unprojected = st_transform(counties, crs = st_crs(states))
stopifnot(st_crs(states) == st_crs(counties_unprojected))
tm_shape(states) + tm_polygons(col = "lightgrey") +
  tm_shape(counties_unprojected) + tm_polygons(col = "darkgreen")
```
