---
title: "My Incredible HTML File"
author: "Nick Hagerty"
date: "9/7/2023"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Biggest header

## Less big header

### Smaller header

#### Small header

**Bold** text. *Italic* text. Link to [Google](http://www.google.com).


Bulleted list of things:

* One thing
* Another thing
  + Sub-thing
  + Sub-thing
* A final thing


Numbered list:

1. Get up
1. Shower
1. Eat breakfast
   a. Drink coffee too
1. Get out the door


Here is a **code chunk** with the default options. It prints both the code and the output:

```{r}
2+2

```

If you want to hide the code, but print the output, use ` ```{r, echo=FALSE}`.

```{r, echo=FALSE}
2+2

```

If you want to hide both the code and the output, but still run it in the background, use ` ```{r, include=FALSE}`.

```{r, include=FALSE}
2+2

```

If you want to print the code, but not run it, use ` ```{r, eval=FALSE}`.

```{r, eval=FALSE}
2+2

```


A horizontal line follows.

***

## Latex code

You can write math formulas inline like this: $2 \times 3 = 6$ or on their own line like this:

$$y_i = \alpha + \beta x_i + \epsilon_i$$

Easy!

If you don't leave a blank line in between paragraphs, it won't give you a line break.
Unless you put a backslash at the end of the line...\
Then it will!


