---
title: "Topic 13: Prediction Methods"
subtitle: "Part 2: Learning with tidymodels"
author: "Nick Hagerty <br> ECNS 460/560 Fall 2023 <br> Montana State University"
date: "<br> .smallest[*Adapted from [“Prediction and machine-learning in econometrics”](https://github.com/edrubin/EC524W21) by Ed Rubin, used with permission. Slides 1-38 are excluded from this resource's overall CC license.]"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'css/my-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
---
name: toc

```{css, echo=FALSE}
.scroll-output-full {
  height: 90%;
  overflow-y: scroll;
}
.scroll-output-75 {
  height: 75%;
  overflow-y: scroll;
}
```

```{r, setup, include = F}
library(knitr)
library(DT)
# Define colors
red_pink   = "#e64173"
turquoise  = "#20B2AA"
orange     = "#FFA500"
red        = "#fb6107"
blue       = "#3b3b9a"
green      = "#8bb174"
grey_light = "grey70"
grey_mid   = "grey50"
grey_dark  = "grey20"
purple     = "#6A5ACD"
slate      = "#314f4f"
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  cache = T,
  warning = F,
  message = F
)
opts_chunk$set(dev = "svg")
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
options(knitr.table.format = "html")
```


# Table of contents

1. [Setup and splitting](#tidymodels)

1. [Recipes](#recipes)

1. [Workflows](#workflows)

1. [Tuning](#tuning)

1. [Prediction](#prediction)

1. [Dependence](#dependence)


---
layout: true
# Setting up tidymodels

---
class: inverse, middle
name: tidymodels

---

`tidymodels` is a set of packages that makes it easy to set up, estimate, tune, and use predictive models.
- There are many other approaches; this is just one.

Load the following packages:
```{r}
library(pacman)
p_load(
  tidyverse, modeldata, skimr, janitor, summarytools,
  tidymodels, magrittr, glmnet, ISLR
)
```

---

Let's try to **predict people's credit scores,** using other information about them and their credit cards. We'll work with the `Credit` dataset from the `ISLR` package.

.scroll-output-75[
  .smallest[
```{r}
# Load the credit dataset
data(Credit)
# Take a look at the data
skim(Credit)
```
  ]
]

---
layout: true
# Split the data

---

The first thing we need to do is set aside a (randomly selected) subset of our data for testing.
* Remember, **testing is separate from (cross-)validation.** We only look at the test set once we're happy with our predictive model.

We can use some functions from the `rsample` package within `tidymodels`:
```{r}
# Set a seed (ensures reproducible results)
set.seed(12345)
# Create an 80/20 split by random sampling
credit_split = Credit |> initial_split(prop = 0.8)
# Grab each subset
credit_train = credit_split |> training()
credit_test  = credit_split |> testing()
credit_split
```


---
layout: true
# Recipes

---
name: recipes
class: inverse, middle

---

One advantage of `tidymodels` is that it makes it really easy to **clean** a lot of variables at once.

Examples of operations in which this might be useful:
* standardize *all* your numeric variables
* create dummies for *all* your categorical variables
* remove variables that are perfectly determined by other variables
* impute missing values

---

First, we **define the recipe.** This step is like listing the ingredients.
* The syntax is like `lm()`, but we aren't estimating anything yet.
* We're going to predict `Rating`, so it goes on the left-hand side.
* `.` means "all other variables".
* No transformations yet - we're just defining the roles of each variable.

```{r}
recipe_all = recipe(Rating ~ ., data = credit_train)
recipe_all
```

---

.smaller[
We now can **add cleaning steps** to the recipe to pre-process our data. The tidymodels galaxy has [many](https://recipes.tidymodels.org/reference/index.html) possible steps for
* imputation: `step_impute_mean()`, `step_impute_mode()`
* transformation: `step_log()`, `step_poly()`, `step_mutate()`
* dummying and discretization: `step_dummy()`, `step_discretize()`
* and many other things: `step_center()`, `step_normalize()`, `step_lag()`

To apply any of these steps, you just need to tell the `step_*()` function which variables you want it to target:
* All: `all_vars()`
* Role: `all_predictors()` or `all_outcomes()` or `has_role()`
* Type: `all_nominal()` or `all_numeric()`
* Variable names, or selectors (e.g., `starts_with()` or `contains()`)
* Unselect using the minus sign (-).

If you have an "ID" variable, you can give it an "ID" role using `update_role(id_var, new_role = "ID")`.
]

---

**Add cleaning steps:** Let's write a step that creates indicator variables for each of our predictors that are categorical or string variables.

```{r}
recipe_all |> step_dummy(all_nominal_predictors())
```

---

**Prep and juice:** To get the processed data frame, we need to add two more things to our recipe:
* `prep()` does the prep work (estimating means for imputation).
* `juice()` applies the preprocessing to the training dataset.

Now the original categorical variables are replaced by indicators for each of their levels.

```{r, eval=F}
credit_clean = recipe_all |> 
  step_dummy(all_nominal_predictors()) |> 
  prep() |>
  juice()
credit_clean |> skim()
```

---

Let's set up and apply a more complex recipe:

.small[
```{r}
credit_recipe = 
  # Define the recipe: Rating predicted by all other vars in credit_train
  recipe(Rating ~ ., data = credit_train) |>
  # Define ID variable (so it's not used in the model)
  update_role(ID, new_role = "ID") |>
  # Impute missing values (use means) for numeric predictors
  step_impute_mean(all_numeric_predictors()) |> 
  # Impute missing values (by k-nearest neighbors) for categorical predictors
  step_impute_knn(all_nominal_predictors(), neighbors = 5) |>
  # Create polynomial terms for numeric predictors
  step_poly(all_numeric_predictors(), degree=2) |>
  # Create indicators for categorical predictors
  step_dummy(all_nominal_predictors()) |>
  # Create interactions (all possible first-degree terms)
  step_interact(~all_predictors():all_predictors()) |>
  # Remove predictors with near-zero variance (improves stability)
  step_nzv(all_predictors())
# Create the cleaned dataset
credit_clean = credit_recipe |> prep() |> juice()
# Skim the cleaned (full) dataset
credit_clean |> skim()
```
]

---
layout: true
# Workflows

---
name: workflows
class: inverse, middle

---

After preprocessing, the next step is to **define a model.** This includes:
* The **model specification** (the general type of model).
* The **engine** (the specific function/package).

Here's a simple linear regression using `lm()`:

```{r}
model_lm = 
  linear_reg() |>
  set_engine("lm")
```

---

Last, we want to **set up any resampling** methods we're going to apply.

Define folds for 5-fold cross-validation:

```{r}
credit_cv = credit_train |> vfold_cv(v = 5)
tidy(credit_cv)
```

---

Now we can **put it all together** with `workflow()`.

```{r}
fit_lm_cv =
  workflow() |>                 # Define a workflow
  add_model(model_lm) |>        # Choose the model
  add_recipe(credit_recipe) |>  # Clean the data
  fit_resamples(credit_cv)       # Estimate & cross-validate
# Check the performance
fit_lm_cv |> collect_metrics()
```

---

To see the performance within each fold:

```{r}
fit_lm_cv |> collect_metrics(summarize = F)
```


---
layout: false
class: inverse, middle
# Tuning

---
layout: true
# Lasso regression example

---
name: tuning

In lasso regression, we don't just want to estimate one model, we also want to **tune** the penalty $\lambda$. That means:
- Estimate the same regression for different values of $\lambda$.
- Choose the one that performs best in cross-validation.

Remember, we also need to standardize all our predictors beforehand.
* Lucky for us, R's `glmnet` engine standardizes everything automatically by default.

---

The first change to make is the **model:**

```{r}
model_lasso = 
  linear_reg(penalty = tune(), mixture = 1) |> 
  set_engine("glmnet")
```

- `penalty = tune()` tells R what parameter we want to tune.
- Set `mixture = 1` means $\alpha=1$ (for ridge, set `mixture = 0`).
- Use the `glmnet` engine.

---

**Then,** instead of using `fit_resamples()`, we're going to `tune_grid()`.

For that we need to define a set of $\lambda$'s to try, spanning several orders of magnitude. Try this:

```{r}
lambdas = 10 ^ seq(from = 5, to = -2, length = 1e3)
```

--

**Now** we can calculate cross-validated MSE for each value of $\lambda$:

```{r}
workflow_lasso = workflow() |>
  add_model(model_lasso) |>
  add_recipe(credit_recipe)
lasso_cv = workflow_lasso |>
  tune_grid(
    credit_cv,
    grid = data.frame(penalty = lambdas),
    metrics = metric_set(rmse)
  )
```

---

Which values of $\lambda$ performed best?

```{r}
lasso_cv |> show_best()
```

---

Let's plot the results for all values of $\lambda$:

```{r, fig.height=4.25, fig.width=8}
autoplot(lasso_cv, metric = "rmse")
```
```{r eval=FALSE, include=FALSE}
# Unused code to plot coefficients by value of lambda
coefs = workflow_lasso |>
  fit(credit_train) |>
  extract_fit_engine() |>
  tidy()
coefs
coefs |>
  ggplot(aes(x=lambda, y=estimate, color=term)) +
    geom_line(show.legend=FALSE) +
    scale_x_log10()
```

---
layout: false
# Ridge example

What if we want to use **ridge** instead of lasso? Just switch `mixture = 0`.

```{r, eval=F}
model_ridge = 
  linear_reg(penalty = tune(), mixture = 0) |> 
  set_engine("glmnet")
workflow_ridge = workflow() |>
  add_model(model_ridge) |>
  add_recipe(credit_recipe2)
credit_cv = credit_train |> vfold_cv(v = 5)
ridge_cv = workflow_ridge |>
  tune_grid(
    credit_cv,
    grid = data.frame(penalty = lambdas),
    metrics = metric_set(rmse)
  )
autoplot(ridge_cv, metric = "rmse")
ridge_cv |> show_best()
```

---
# Elasticnet example

We can use `tune()` to cross validate both $\alpha$ and $\lambda$.

.note[Note] You need to consider all combinations of the two parameters.
<br>This combination can create *a lot* of models to estimate.

1,000 values of $\lambda$ $\times$ 1,000 values of $\alpha$ =
- 1,000,000 models to estimate
- 5,000,000 if you're doing 5-fold CV

To keep your computer from going forever, use a **coarser grid** (i.e., try fewer values).

---
layout: false
class: clear, middle

**Cross-validating elasticnet:**

```{r, credit-net-ex, eval = F}
# Ranges of λ and α
lambdas = 10 ^ seq(from = 5, to = -2, length = 1e2)
alphas = seq(from = 0, to = 1, by = 0.1)

# Define the elasticnet model
model_net = linear_reg(penalty = tune(), mixture = tune()) |>
  set_engine("glmnet")

# Define the workflow
workflow_net = workflow() |>
  add_recipe(credit_recipe) |>
  add_model(model_net)

# CV elasticnet with our ranges of lambdas and alphas
cv_net = 
  workflow_net |>
  tune_grid(
    credit_cv,
    grid = expand_grid(mixture = alphas, penalty = lambdas),
    metrics = metric_set(rmse)
  )
```

---
layout: false
class: clear, middle

**Alternative approach:** `grid_regular()` automatically chooses sensible values of your parameters to try.

```{r, credit-net-ex2, eval = F}
# Define the elasticnet model
model_net = linear_reg(penalty = tune(), mixture = tune()) |>
  set_engine("glmnet")

# Define the workflow
workflow_net = workflow() |>
  add_recipe(credit_recipe) |>
  add_model(model_net)

# CV elasticnet with grid_regular()
cv_net = 
  workflow_net |>
  tune_grid(
    credit_cv,
    grid = grid_regular(mixture(), penalty(), levels=5:5), #<<
    metrics = metric_set(rmse)
  )
```


---
layout: true
# Prediction

---
name: prediction
class: inverse, middle

---

After tuning the penalty through cross-validation, we need to:
1. **Tell** R which model we're choosing to carry forward.
1. **Fit** the model on the *entire* training dataset (not just 4 of 5 folds).
1. **Make predictions** in the testing dataset.

---

Tell R which model we're choosing:

.smaller[
```{r}
final_lasso = 
  workflow_lasso |>
  finalize_workflow(select_best(lasso_cv, metric = "rmse"))
final_lasso
```
]

---

Fit the model and make predictions in the test set:

```{r}
final_fit_lasso = final_lasso |> last_fit(credit_split)
final_fit_lasso |> collect_metrics()
```

---

.smaller[

Now let's take a look at the coefficients from the best model:

```{r}
coefs = final_fit_lasso |>
  extract_fit_parsnip() |>
  tidy()
```

```{R, echo = F}
datatable(coefs, options = list(dom = 'tp'))
```

]

---

How about listing only the non-zero coefficients?

```{r}
coefs_nonzero = coefs |>
  filter(estimate > 0)
```

```{R, echo = F}
datatable(coefs_nonzero, options = list(dom = 'tp'))
```

---

Finally, we can plot the predictions against the true values:

```{r, fig.height=3.75, fig.width=8}
final_fit_lasso |> collect_predictions() |>
  ggplot(aes(x=.pred, y=Rating)) + 
  geom_abline(slope=1) +
  geom_point(color="blue")
```

---
layout: true
# Dependence

---
class: inverse, middle

---
name: dependence

.b[Resampling methods] assume something similar to independence: our resampling must match the original sampling procedure.

If observations are "linked" but we resample independently, CV may break.

If we have .hi-pink[repeated observations] on individuals $i$ through time $t$ (i.e., time series or panel data):
- It's pretty likely $y_{i,t}$ and $y_{i,t+1}$ are related—and maybe $y_{i,t+\ell}$.
- Initial sample may draw individuals $i$, but .it[standard] CV ignores time.

In other case, .hi-pink[some individuals are linked with other individuals], *e.g.*,
- $y_{i,t}$ and $y_{j,t}$ my be correlated if $i$ and $j$ live togother
- Also: $y_{i,t}$ and $y_{j,t+\ell}$ could be correlated

---

In other words: Spatial or temporal dependence between observations .b[breaks the separation between training and testing samples].

Breaking this separation train-test separation leads us back to

- .b[Overfitting] the sample—since training and testing samples are linked

- .b[Overestimating model performance]—the estimated test MSE will be more of a training MSE.

.b[Solutions] to this problem involve .hi-pink[matching the resampling process] to the .hi-pink[original sampling and underlying dependence].

---
## 1. Cross-sectional spatial data

.small[
Observations near each other in space are probably related to each other. We may want to resample blocks of larger areas, not individual observations.

- **Training/testing split?** Use `group_initial_split()` instead of `initial_split()` to keep groups of observations together when splitting the data.

- **Cross-validation:** Use `group_vfold_cv()` instead of `vfold_cv()` to resample by grabbing entire blocks at once (e.g., all counties for a given state, instead of treating each county as independent). Example:
]
```{r, eval=F}
# 5-fold CV grabbing entire states at once
cv_block = df_train |> group_vfold_cv(group=state, v=5)
```

---
## 2. Panel data

.small[
Repeated observations for the same unit (person, county, etc.) are related to each other. We want to resample entire units at a time, not allow data from the same unit to show up in both training & testing data. 

- **Training/testing split?** Use `group_initial_split()`, setting the `group` argument to the unit ID variable.

- **Cross-validation:** Use `group_vfold_cv()`, setting the `group` argument to the unit ID variable.

*Time series or panel?* Let $N=$ number of units, $T=$ number of time periods. Generally, if $N>T$ you have panel data; if $T>N$ you have time series data.
]

---
## 3. Time series data

.small[
Later observations may "remember" (contain information about) earlier observations. We need to train *only* on data that comes *before* the testing or validation set.

- **Training/testing split:** Use `initial_time_split()` instead of `initial_split()`. This makes your training data come first in time, and your training data last, instead of random sampling.

- **Cross-validation:** Use `sliding_index()` (or related functions [here](https://rsample.tidymodels.org/reference/slide-resampling.html)) instead of `vfold_cv()` to resample by choosing points at which we will train in the "past" and validate in the "future" ([more info](https://medium.com/@soumyachess1496/cross-validation-in-time-series-566ae4981ce4)). Example:
]
```{r, eval=F}
# This generates folds with 7 days for training, 1 day for assessment.
cv_time = df_train |> sliding_index(index=DATE_VAR, lookback=7, assess_stop=1)
```



