---
title: "Lecture 14a: Shrinkage Methods"
author: "Nick Hagerty <br> ECNS 491/560 Fall 2022 <br> Montana State University"
date: "<br> *Adapted with permission from materials by [Ed Rubin](https://github.com/edrubin/EC524W21)."
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'css/my-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
---
exclude: true

```{r, setup, include = F}
library(pacman)
p_load(
  broom, tidyverse,
  ggplot2, ggthemes, ggforce, ggridges, cowplot, scales,
  latex2exp, viridis, extrafont, gridExtra, plotly, ggformula,
  kableExtra, DT,
  snakecase, janitor, knitr, future, furrr,
  MASS, estimatr, tidymodels, glmnet,
  huxtable, here, magrittr, parallel
)
# Define colors
red_pink   = "#e64173"
turquoise  = "#20B2AA"
orange     = "#FFA500"
red        = "#fb6107"
blue       = "#3b3b9a"
green      = "#8bb174"
grey_light = "grey70"
grey_mid   = "grey50"
grey_dark  = "grey20"
purple     = "#6A5ACD"
slate      = "#314f4f"
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = F,
  message = F
)
opts_chunk$set(dev = "svg")
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
options(knitr.table.format = "html")
```

---

# Table of contents

1. [Shrinkage methods](#intro)

1. [Ridge regression](#ridge)

1. [Lasso (and elasticnet)](#lasso)


---
layout: true
# Shrinkage methods

---
name: intro
## Intro

Linear regression is great for causal inference.
* Best linear approximation to the CEF.
* Best linear unbiased estimator (BLUE) (under certain assumptions).
* Can give us unbiased treatment effects (under certain assumptions).

Is it good for prediction too?
* More advanced, nonlinear methods sometimes work better.
* But linear regression can still work well even on fairly complex problems.

---

To use regression effectively in prediction, we have to deal with two issues:

**Model selection.** How do we choose the $X_i$'s?
- Which variables do you put in your regression?
- How do you parameterize/transform/interact them?
- What if you have more variables than observations?
- More variables improve fit, but how do you avoid overfitting?

**Variance reduction.** Can we improve predictions with some tweaks?
* OLS tends to have low bias.super[.pink[†]] and relatively low variance.
* But if we accept a little more bias, we can get big reductions in variance, and therefore better predictions (lower MSE).

.footnote[
.pink[†] Unbiased, in fact, when the true DGP (population CEF) is linear.
]

---

## Shrinkage methods
- fit a model that contains .pink[all] $\color{#e64173}{p}$ .pink[predictors]
- simultaneously: .pink[**shrink** coefficients] toward zero
  - Shrink: constrain, regularize.

.note[Idea:] Penalize the model for coefficients as they move away from zero.

---
## Intuition

When we estimate a large, imprecise coefficient, should we just accept it?
- Is the pattern likely to come up again, or idiosyncratic to this sample?
- Maybe we should demand a higher bar for larger coefficients.

```{r jabba, echo=FALSE, fig.height=3.6, fig.width=7}
ggplot(starwars %>% filter(height>140, height<180) %>% slice(1:16), aes(x=height, y=mass)) +
  geom_point() +
  geom_smooth(method="lm", formula=y~x)
```

---
name: shrinkage-why
## Overall idea

- Shrinking our coefficients toward zero .hi[reduces the model's variance]..super[.pink[†]]
- .hi[Penalizing] our model for .hi[larger coefficients] shrinks them toward zero.
- The .hi[optimal penalty] will balance reduced variance with increased bias.

.footnote[
.pink[†] Imagine the extreme case: a model whose coefficients are all zeros has no variance.
]

--

Now you understand shrinkage methods.
- .attn[Ridge regression]
- .attn[Lasso]
- .attn[Elasticnet]

---
layout: true
# Ridge regression

---
class: inverse, middle

---
name: ridge
## Back to least squares

.note[Recall] Least-squares regression gets $\hat{\beta}_j$'s by minimizing RSS, _i.e._,
$$
\begin{align}
  \min_{\hat{\beta}} \text{RSS} = \min_{\hat{\beta}} \sum_{i=1}^{n} e_i^2 = \min_{\hat{\beta}} \sum_{i=1}^{n} \bigg( \color{#FFA500}{y_i} - \color{#6A5ACD}{\underbrace{\left[ \hat{\beta}_0 + \hat{\beta}_1 x_{i,1} + \cdots + \hat{\beta}_p x_{i,p} \right]}_{=\hat{y}_i}} \bigg)^2
\end{align}
$$

--

.attn[Ridge regression] makes a small change
- .pink[adds a shrinkage penalty] = the sum of squared coefficients $\left( \color{#e64173}{\lambda\sum_{j}\beta_j^2} \right)$
- .pink[minimizes] the (weighted) sum of .pink[RSS and the shrinkage penalty]

--

$$
\begin{align}
  \min_{\hat{\beta}^R} \sum_{i=1}^{n} \bigg( \color{#FFA500}{y_i} - \color{#6A5ACD}{\hat{y}_i} \bigg)^2 + \color{#e64173}{\lambda \sum_{j=1}^{p} \beta_j^2}
\end{align}
$$

---
name: ridge-penalization

.col-left[
.hi[Ridge regression]
$$
\begin{align}
\min_{\hat{\beta}^R} \sum_{i=1}^{n} \bigg( \color{#FFA500}{y_i} - \color{#6A5ACD}{\hat{y}_i} \bigg)^2 + \color{#e64173}{\lambda \sum_{j=1}^{p} \beta_j^2}
\end{align}
$$
]

.col-right[
.b[Least squares]
$$
\begin{align}
\min_{\hat{\beta}} \sum_{i=1}^{n} \bigg( \color{#FFA500}{y_i} - \color{#6A5ACD}{\hat{y}_i} \bigg)^2
\end{align}
$$
]

<br><br><br><br>

$\color{#e64173}{\lambda}\enspace (\geq0)$ is a tuning parameter for the harshness of the penalty.
<br>
$\color{#e64173}{\lambda} = 0$ implies no penalty: we are back to least squares.
--
<br>
Each value of $\color{#e64173}{\lambda}$ produces a new set of coefficients.

--

Ridge's approach to the bias-variance tradeoff: Balance
- reducing .b[RSS], _i.e._, $\sum_i\left( \color{#FFA500}{y_i} - \color{#6A5ACD}{\hat{y}_i} \right)^2$
- reducing .b[coefficients] .grey-light[(ignoring the intercept)]

$\color{#e64173}{\lambda}$ determines how much ridge "cares about" these two quantities..super[.pink[†]]

.footnote[
.pink[†] With $\lambda=0$, least-squares regression only "cares about" RSS.
]
---
layout: false
class: clear, middle

.b[Ridge regression coefficients] for $\lambda$ between 0.01 and 100,000
```{r, plot-ridge-glmnet, echo = F}
# Load the credit dataset
credit_df = ISLR::Credit %>% clean_names()
# Processing recipe: Define ID, standardize, create dummies, rename (lowercase)
credit_recipe = credit_df %>% recipe(balance ~ .) %>% 
  update_role(id, new_role = "id variable") %>% 
  step_normalize(all_predictors() & all_numeric()) %>% 
  step_dummy(all_predictors() & all_nominal()) %>% 
  step_rename_at(everything(), fn = str_to_lower)
# Time to juice
credit_clean = credit_recipe %>% prep() %>% juice()

# Define our range of lambdas (glmnet wants decreasing range)
lambdas = 10^seq(from = 5, to = -2, length = 100)
# Fit ridge regression
est_ridge = glmnet(
  x = credit_clean %>% dplyr::select(-balance, -id) %>% as.matrix(),
  y = credit_clean$balance,
  standardize = F,
  alpha = 0,
  lambda = lambdas
)

ridge_df = est_ridge %>% coef() %>% t() %>% as.matrix() %>% as.data.frame()
ridge_df %<>% dplyr::select(-1) %>% mutate(lambda = est_ridge$lambda)
ridge_df %<>% gather(key = "variable", value = "coefficient", -lambda)
ggplot(
  data = ridge_df,
  aes(x = lambda, y = coefficient, color = variable)
) +
geom_line() +
scale_x_continuous(
  expression(lambda),
  labels = c("0.1", "10", "1,000", "100,000"),
  breaks = c(0.1, 10, 1000, 100000),
  trans = "log10"
) +
scale_y_continuous("Ridge coefficient") +
scale_color_viridis_d("Predictor", option = "magma", end = 0.9) +
theme_minimal(base_size = 20, base_family = "Fira Sans Book") +
theme(legend.position = "bottom")

```

---
layout: true
# Ridge regression

---

## $\lambda$ and penalization

Choosing a .it[good] value for $\lambda$ is key.
- If $\lambda$ is too small, then our model is essentially back to OLS.
- If $\lambda$ is too large, then we shrink all of our coefficients too close to zero.

--

.qa[Q] So what do we do?
--
<br>
.qa[A] Cross validate!

---
## Penalization

.note[Note] Because we sum the .b[squared] coefficients, we penalize increasing .it[big] coefficients much more than increasing .it[small] coefficients.

.ex[Example] For a value of $\beta$, we pay a penalty of $2 \lambda \beta$ for a small increase..super[.pink[†]]

.footnote[
.pink[†] This quantity comes from taking the derivative of $\lambda \beta^2$ with respect to $\beta$.
]

- At $\beta = 0$, the penalty for a small increase is $0$.
- At $\beta = 1$, the penalty for a small increase is $2\lambda$.
- At $\beta = 2$, the penalty for a small increase is $4\lambda$.
- At $\beta = 3$, the penalty for a small increase is $6\lambda$.
- At $\beta = 10$, the penalty for a small increase is $20\lambda$.

Now you see why we call it .it[shrinkage]: it encourages small coefficients.
---
name: standardization
## Penalization and standardization

.attn[Important] Predictors' .hi[units] can drastically .hi[affect ridge regression results].

.b[Why?]
--
 Because $\mathbf{x}_j$'s units affect $\beta_j$, and ridge is very sensitive to $\beta_j$.

--

.ex[Example] Let $x_1$ denote distance.

.b[Least-squares regression]
<br>
If $x_1$ is .it[meters] and $\beta_1 = 3$, then when $x_1$ is .it[km], $\beta_1 = 3,000$.
<br>
The scale/units of predictors do not affect least squares' estimates.

--

.hi[Ridge regression] pays a much larger penalty for $\beta_1=3,000$ than $\beta_1=3$.
<br>You will not get the same (scaled) estimates when you change units.

--

.note[Solution] Standardize your variables, _i.e._, `x_stnd = (x - mean(x))/sd(x)`.







---
layout: true
# Lasso

---
class: inverse, middle
---
name: lasso
## Intro

.attn[Lasso] simply replaces ridge's .it[squared] coefficients with absolute values.

--

.hi[Ridge regression]
$$
\begin{align}
\min_{\hat{\beta}^R} \sum_{i=1}^{n} \big( \color{#FFA500}{y_i} - \color{#6A5ACD}{\hat{y}_i} \big)^2 + \color{#e64173}{\lambda \sum_{j=1}^{p} \beta_j^2}
\end{align}
$$

.hi-grey[Lasso]
$$
\begin{align}
\min_{\hat{\beta}^L} \sum_{i=1}^{n} \big( \color{#FFA500}{y_i} - \color{#6A5ACD}{\hat{y}_i} \big)^2 + \color{#8AA19E}{\lambda \sum_{j=1}^{p} \big|\beta_j\big|}
\end{align}
$$

Everything else will be the same—except one aspect...

---
name: lasso-shrinkage
## Shrinkage

Unlike ridge, lasso's penalty does not increase with the size of $\beta_j$.

You always pay $\color{#8AA19E}{\lambda}$ to increase $\big|\beta_j\big|$ by one unit.

--

The only way to avoid lasso's penalty is to .hi[set coefficients to zero].

--

This feature has two .hi-slate[benefits]
1. Some coefficients will be .hi[set to zero]—we get "sparse" models.
1. Lasso can be used for subset/feature .hi[selection].

--

We will still need to carefully select $\color{#8AA19E}{\lambda}$.

---
layout: false
class: clear, middle

.b[Lasso coefficients] for $\lambda$ between 0.01 and 100,000
```{r, plot-lasso-glmnet, echo = F}

# Define our range of lambdas (glmnet wants decreasing range)
lambdas = 10^seq(from = 5, to = -2, length = 100)
# Fit lasso regression
est_lasso = glmnet(
  x = credit_clean %>% dplyr::select(-balance, -id) %>% as.matrix(),
  y = credit_clean$balance,
  standardize = F,
  alpha = 1,
  lambda = lambdas
)

lasso_df = est_lasso %>% coef() %>% t() %>% as.matrix() %>% as.data.frame()
lasso_df %<>% dplyr::select(-1) %>% mutate(lambda = est_lasso$lambda)
lasso_df %<>% gather(key = "variable", value = "coefficient", -lambda)
ggplot(
  data = lasso_df,
  aes(x = lambda, y = coefficient, color = variable)
) +
geom_line() +
scale_x_continuous(
  expression(lambda),
  labels = c("0.1", "10", "1,000", "100,000"),
  breaks = c(0.1, 10, 1000, 100000),
  trans = "log10"
) +
scale_y_continuous("Lasso coefficient") +
scale_color_viridis_d("Predictor", option = "magma", end = 0.9) +
theme_minimal(base_size = 20, base_family = "Fira Sans Book") +
theme(legend.position = "bottom")
```

---
class: clear, middle
layout: false

So which shrinkage method should you choose?

---
layout: true
# Ridge or lasso?

---
name: or

.col-left.pink[
.b[Ridge regression]
<br>
<br>.b.orange[+] shrinks $\hat{\beta}_j$ .it[near] 0
<br>.b.orange[-] many small $\hat\beta_j$
<br>.b.orange[-] doesn't work for selection
<br>.b.orange[-] difficult to interpret output
<br>.b.orange[+] better when all $\beta_j\neq$ 0
]

.col-right.purple[
.b[Lasso]
<br>
<br>.b.orange[+] shrinks $\hat{\beta}_j$ to 0
<br>.b.orange[+] many $\hat\beta_j=$ 0
<br>.b.orange[+] great for selection
<br>.b.orange[+] sparse models easier to interpret
<br>.b.orange[-] implicitly assumes some $\beta=$ 0
]

.left-full[
> [N]either ridge... nor the lasso will universally dominate the other.

.ex[ISL, p. 224]
]

---
name: both
layout: false
# Ridge .it[and] lasso
## Why not both?

.hi-blue[Elasticnet] combines .pink[ridge regression] and .grey[lasso].

--

$$
\begin{align}
\min_{\beta^E} \sum_{i=1}^{n} \big( \color{#FFA500}{y_i} - \color{#6A5ACD}{\hat{y}_i} \big)^2 + \color{#181485}{(1-\alpha)} \color{#e64173}{\lambda \sum_{j=1}^{p} \beta_j^2} + \color{#181485}{\alpha} \color{#8AA19E}{\lambda \sum_{j=1}^{p} \big|\beta_j\big|}
\end{align}
$$

We now have two tuning parameters: $\lambda$ (penalty) and $\color{#181485}{\alpha}$ (mixture).

- $\color{#e64173}{\alpha = 0}$ specifies ridge
- $\color{#8AA19E}{\alpha=1}$ specifies lasso


